Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-26 paper/geometry-twocolumn.tex
Patty's fixes to Section 2

This approach regards the difference between points and vectors as a
change of data representation. For example, if $(0,0)$ and $(10,20)$
are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ the
tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
  point} because they have the same displacement from these two
origins, respectively.
%It is merely an artifact of the representation of points as pairs of
%numbers that they appear to be different.
This suggests that programs that manipulate points %data representing points
should be invariant with respect to changes of
origin. Programs that manipulate vectors, on the other hand, should
not be invariant under change of origin. Different vectors represent
different offsets, and the vector $(0,0)$ always represents the zero
offset.

@@

Reynolds' relational approach can be applied in the geometric setting
to show how quantifying over all changes of origin ensures the
invariance of programs under any particular choice of origin. For
this, we first define a family of binary relations on $\mathbb{R}^2$
that is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group
$\Transl{2}$ of translations under addition. The $\Transl{2}$-indexed
family of binary relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2
\times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$ is then defined by
%\begin{displaymath}
$R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}$.
%\end{displaymath}
We then consider a function $f$ that takes as input two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that the tuples all represent points with respect to the same origin,
and that $f$ is invariant under the choice of origin.  Reynolds'
relational approach formalises this intention precisely: for any
$\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
%$f$
%should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}},
  (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}),
  f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation:
%\statementref{eq:f-preserve-rel-frame} is equivalent to: 
%for all $\vec{t} \in \Transl{2}$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} +
  \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
Thus, Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

@@

Reynolds also showed how a type discipline can be used to establish
that (the denotational interpretations of) programs preserve
relations. For Reynolds, the type discipline of interest was that of
the polymorphic $\lambda$-calculus, which supports the construction of
new types by universal quantification over types.
%, in which new types can be
%constructed by universal quantification over all types.
In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. By contrast, in our statements of geometric invariance in
\autoref{sec:type-system-geom-intro} we did not quantify over all
relations, but instead quantified over all changes of origin and used
a specific choice of origin to select a relation from the family
$\{R_{\vec{t}}\}$. This suggests introducing quantification over
changes of origin into the language of types. We use the notation
$\forall t \mathord: \SynTransl{2}. A$ for quantification over all
2-dimensional translations (i.e.,~choices of origin) $t$, and refer to
$\SynTransl{2}$ as the \emph{sort} of $t$. Note the difference in
fonts used to distinguish the semantic group $\Transl{2}$ from the
syntactic sort $\SynTransl{2}$. We use a similar convention
below, too.

@@

At the end of \autoref{sec:oiri} we considered functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that preserve all
changes of origin. This property of $f$ can be expressed in terms of
types by
%\begin{displaymath}
$  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
  \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
%\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above and the standard relational interpretations for
tuple and function types, we recover
\statementref{eq:f-invariant-frame} exactly.

@@

Although vector addition, negation, and scaling are not invariant
under change of origin, they are invariant under change of basis. As
with origin invariance, we can express basis invariance as
preservation of relations indexed by changes of basis. Change of basis
is achieved by applying an invertible linear map, and the collection
of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
group $\GL{2}$ which we represent in our language by a new indexing
sort $\SynGL{2}$ with (non-abelian) group structure that we will write
multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and relational
semantics of $\tyPrim{vec}{B,t}$ are given by:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

>>>>

This approach regards the difference between points and vectors as a
change of data representation. For example, if $(0,0)$ and $(10,20)$
are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ and the
tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
  point} because they have the same displacement from these two
origins, respectively.
%It is merely an artifact of the representation of points as pairs of
%numbers that they appear to be different.
This suggests that programs that manipulate points %data representing points
should be invariant with respect to changes of
origin. Programs that manipulate vectors, on the other hand, should
not be invariant under change of origin. Different vectors represent
different offsets, and the vector $(0,0)$ always represents the zero
offset.

@@

Reynolds' relational approach can be applied in the geometric setting
to show how quantifying over all changes of origin ensures the
invariance of programs under any particular choice of origin. For
this, we first define a family of binary relations on $\mathbb{R}^2$
that is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group
$\Transl{2}$ of translations under addition. The $\Transl{2}$-indexed
family of binary relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2
\times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$ is then defined by
%\begin{displaymath}
$R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}$.
%\end{displaymath}
We then consider a function $f$ that takes as input two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that the tuples all represent points with respect to the same origin,
and that $f$ is invariant under the choice of origin.  Reynolds'
relational approach formalises this intention precisely. For any
$\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
%$f$
%should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}},
  (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}),
  f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation,
%\statementref{eq:f-preserve-rel-frame} is equivalent to: 
again for all $\vec{t} \in \Transl{2}$:
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} +
  \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
Thus, Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

@@

Reynolds also showed how a type discipline can be used to establish
that (the denotational interpretations of) programs preserve
relations. For Reynolds, the type discipline of interest was that of
the polymorphic $\lambda$-calculus, which supports the construction of
new types by universal quantification over types.
%, in which new types can be
%constructed by universal quantification over all types.
In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. By contrast, in our statements of geometric invariance in
\autoref{sec:oiri} we did not quantify over all
relations, but instead quantified over all changes of origin and used
a specific choice of origin to select a relation from the family
$\{R_{\vec{t}}\}$. This suggests introducing quantification over
changes of origin into the language of types. We use the notation
$\forall t \mathord: \SynTransl{2}. A$ for quantification over all
2-dimensional translations (i.e.,~choices of origin) $t$, and refer to
$\SynTransl{2}$ as the \emph{sort} of $t$. Note the difference in
fonts used to distinguish the semantic group $\Transl{2}$ from the
syntactic sort $\SynTransl{2}$. We use a similar convention
below, too.

@@

At the end of \autoref{sec:oiri} we considered functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that preserve all
changes of origin. This property of $f$ can be expressed in terms of
types by
%\begin{displaymath}
$  \mathit{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
  \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
%\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above and the standard relational interpretations for
tuple and function types, we recover
\statementref{eq:f-invariant-frame} exactly.

@@

Although vector addition, negation, and scaling are not invariant
under change of origin, they are invariant under change of basis. As
with origin invariance, we can express basis invariance as
preservation of relations indexed by changes of basis. Change of basis
is achieved by applying an invertible linear map, and the collection
of all such maps on $\mathbb{R}^2$ forms the \emph{General Linear}
group $\GL{2}$, which we represent in our language by a new indexing
sort $\SynGL{2}$ with (non-abelian) group structure that we will write
multiplicatively. We then extend $\tyPrimNm{vec}$ to allow indices of
sort $\SynGL{2}$, as well as $\SynTransl{2}$, so that
$\tyPrim{vec}{B,t}$ is a vector that varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and relational
semantics of $\tyPrim{vec}{B,t}$ are given by:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'})
    \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}
